From DSL to HPC Component-Based Runtime:\\ A Multi-Stencil DSL Case Study

### 2015-08-08 WOLFHPC15/article/evaluation.tex
fix all helene 08_08

The series-parallel tree decomposition extracted by the MSC transformation is composed of seventeen sequences and eighteen parallel sections. Figure~\ref{fig:freq} represents for a given level of parallelism, \ie the number of tasks which can be executed concurrently, the number of time this level is observed in the final component assembly. One can notive that the level of task parallelism extracted from the Shallow water equations is limited by at the two sequential parts of the application (level 1). However, as a level of sixteen parallel tasks is reach too times, but also five times for the level six, sequential restrictions could be amortize. Moreover, as the level of parallelism in the application is heterogenous, the choice of the number of threads to launch for task parallelism is difficult.

@@

Figure~\ref{fig:exectime} illustrates the execution time for each step of the MSC transformation for an overall execution time of ten seconds. Execution times have been computed on a laptop with a bi-core Intel Core i5 1,4 GHz, and 8Gb LPDDR3. 
\begin{figure}[!h]
 \begin{center}
 \begin{tabular}{c|c|c|c|c|}
   step & $\Gamma_{sync}$ & $\Gamma_{dep}$ & $\Gamma_{msp}$ & $\Gamma_{tsp}$\\
   \hline
   time (ms) & 2 & 530 & 8297 & 1133\\
   \hline
   \% & 0.02 & 5.3 & 83.3 & 11.37\\
 \end{tabular}
\caption{Execution times of the MSC transformation steps}
\label{fig:exectime}
 \end{center}
\end{figure}
One can notice that the transformation of $\Gamma_{dep}$ to a minimal series-parallel graph is the longest step of MSC, because of the removal of the forbidden shapes in the graph. Actually, the number of forbidden shapes removed in $\Gamma_{dep}$ is not counted, because the algorithm uses a general solution instead of finding each forbidden shape, but it seems that many of them appeared. The \emph{N-shape}, represented in Figure~\ref{fig:n} is forbidden in a minimal series-parallel graph as it is not possible to exactly express it using sequences and parallel sections.
The fact that many \emph{N-shape} are removed in $\Gamma_{dep}$ shows that the creation of a static shedule of tasks may not be the best solution for complex simulations. Actually, if a \emph{N-shape} cannot be represented by sequences and parallel sections, this shape can perfectly be handled by a dynamic scheduler or by the direct static representation of $\Gamma_{dep}$ where tasks wait for their input before starting.

\subsubsection*{Preliminary Performances}
To evaluate if performances of the overall automatic parallelization are promising we have proceeded as follows. Our current implementation of the MSCAC compiler generates a back-end code using the SkelGIS library. For this reason, our evaluation compare the shallow water equations first parallelized with a pure SkelGIS code (data structures, applicators and interfaces of SkelGIS~\cite{CPE:CPE3494}), and second parallelized with MSL (which uses the SkelGIS data structure). As the SkelGIS library has proved its scalability on the Shallow water equations compared to an MPI parallelization, the evaluation is relevant to evaluate scalability of MSL. 
Moreover, as SkelGIS library handles data parallelization for distributed memory architectures, we have limited for now our evaluations to the fusionned data parallelization of MSCAC (dumped from $\Gamma_{data}$).

\fix{HC : preciser la machine}

Evaluations have been performed on ... . Figure~\ref{fig:times} shows the execution times in seconds as a function of the number of cores. Figure~\ref{fig:speedup} illustrates a speedup comparison, using the minimum sequential reference time (which is the MSL sequential time). Finally, Figure~\ref{fig:log2} illustrates the logarithmic scale of execution times as a function of the logarithmic scale of the number of cores.
\begin{figure}[!h]
 \begin{center}
 \begin{tabular}{c|c|c|c|c|c|c}
   cores & 8 & 16 & 32 & 64 & 128 & 256\\
   \hline
   MSL & 998.10 & 544.86 & 285.73 & 139.8 & 76.45 & 48.05\\
   \hline
   SkelGIS & 1324.9 & 638.5 & 371.62 & 197.32 & 117.05 & 64.96\\
 \end{tabular}
\caption{Execution times for the shallow water equations using MSL and SkelGIS (in seconds).}
\label{fig:times}
 \end{center}
\end{figure}
\begin{figure*}
\begin{center}
\subfloat[\label{fig:speedup}]{
\resizebox{8cm}{!}{\includegraphics{./images/speedup.pdf}}
}
\hspace{10pt}
\subfloat[\label{fig:log2}]{
\resizebox{8cm}{!}{\includegraphics{./images/logtime.pdf}}
}
\end{center}
\caption{(a) Speedup comparison, and (b) logarithmic execution times, both for the Shallow water equations, using pure SkelGIS and MSL with the SkelGIS data structure.}
\label{fig:perfs}
\end{figure*}

One can notive that execution times using MSL, which itself uses the SkelGIS distributed data structures, is improved compared to a pure SkelGIS parallelization. Moreover, the scalability stays close to the scalability of SkelGIS even if the makespan is improved.

>>>>

The series-parallel tree decomposition $\Gamma_{tsp}$ of this simulation, extracted by the MSC transformation, is composed of seventeen sequence nodes and eighteen parallel nodes. Figure~\ref{fig:freq} represents for a given level of parallelism, \ie the number of tasks to perform concurrently, the number of time this level is observed in the final component assembly. One can notice that the level of task parallelism extracted from the Shallow water equations is limited by two sequential parts in the application (level 1). As a level of sixteen parallel tasks is reached too times, and also five times for the level six, sequential restrictions could be amortized. However, it is rather clear that the task parallelization technique should be used to improve the data parallelism when reaching at its limits, but not to directly get performance. Moreover, as the level of parallelism in the application is heterogenous, the number of threads to launch for task parallelism, and the number of cores to keep for data parallelism is difficult to choose.

@@

\begin{figure}[!h]
 \begin{center}
 \begin{tabular}{c|c|c|c|c|}
   step & $\Gamma_{sync}$ & $\Gamma_{dep}$ & $\Gamma_{msp}$ & $\Gamma_{tsp}$\\
   \hline
   time (ms) & 2 & 530 & 8297 & 1133\\
   \hline
   \% & 0.02 & 5.3 & 83.3 & 11.37\\
 \end{tabular}
\caption{Execution times of the MSC transformation steps}
\label{fig:exectime}
 \end{center}
\end{figure}

Figure~\ref{fig:exectime} illustrates the execution time for each step of the MSC transformation for an overall execution time of ten seconds. Execution times have been computed on a laptop with a bi-core Intel Core i5 1,4 GHz, and 8Gb DDR3. 
One can notice that the transformation of $\Gamma_{dep}$ to a minimal series-parallel graph is the longest step of MSC, because of the removal of the forbidden shapes in the graph. Actually, the number of forbidden shapes removed in $\Gamma_{dep}$ is not counted by the algorithm, which applies a general solution (instead of finding each forbidden shape), but it seems that many of them appear. The \emph{N-shape}, represented in Figure~\ref{fig:n} is forbidden in a minimal series-parallel graph as it is not possible to exactly express it using sequences and parallel sections.
The fact that many \emph{N-shape} are removed in $\Gamma_{dep}$ shows that the creation of a static shedule of tasks may not be the best solution for complex simulations. Actually, a \emph{N-shape} could perfectly be handled by a dynamic scheduler for example. It has to be added, however, that the computation of $\Gamma_{tsp}$ is still interesting to detect how space loops can be fusionned in $\Gamma_{data}$.

\subsubsection*{Preliminary Performances}
To evaluate performance of the generated component-based parallel structure, we have proceeded as follows. Our current implementation of the MSCAC compiler generates a back-end code using the SkelGIS library. For this reason, our evaluation compare the shallow water equations first parallelized with a pure SkelGIS code (data structures, applicators and interfaces of SkelGIS~\cite{CPE:CPE3494}), and second parallelized with MSL (which uses the SkelGIS data structure). As the SkelGIS library has proved its scalability on the Shallow water equations compared to an MPI parallelization, this performance evaluation should be relevant. 
Moreover, as SkelGIS library handles data parallelization for distributed memory architectures, we have limited for now our evaluations to the fusionned data parallelization of MSCAC (dumped from $\Gamma_{data}$).

Evaluations have been performed on TGCC Curie \emph{thin nodes}. Each node is equiped with an Intel E5-2680 (Sandy-Bridge EP) processor containing 16 cores, 64GB of memory, and is linked to an Infiniband network. 
\fix{HC: what are the experiments presented ? Size of the mesh, number of time iterations}

Figure~\ref{fig:times} shows the execution times in seconds as a function of the number of cores. Figure~\ref{fig:speedup} illustrates a speedup comparison, using the minimum sequential reference time (which is the MSL sequential time). Finally, Figure~\ref{fig:log2} illustrates the logarithmic scale of execution times as a function of the logarithmic scale of the number of cores.
\begin{figure}[!h]
 \begin{center}
 \begin{tabular}{c|c|c|c|c|c|c}
   Cores & 8 & 16 & 32 & 64 & 128 & 256\\
   \hline
   MSL & 998.10 & 544.86 & 285.73 & 139.8 & 76.45 & 48.05\\
   \hline
   SkelGIS & 1324.9 & 638.5 & 371.62 & 197.32 & 117.05 & 64.96\\
 \end{tabular}
\caption{Execution times for the shallow water equations using MSL and SkelGIS (in seconds).}
\label{fig:times}
 \end{center}
\end{figure}
\begin{figure*}
\begin{center}
\subfloat[\label{fig:speedup}]{
\resizebox{8cm}{!}{\includegraphics{./images/speedup.pdf}}
}
\hspace{10pt}
\subfloat[\label{fig:log2}]{
\resizebox{8cm}{!}{\includegraphics{./images/logtime.pdf}}
}
\end{center}
\caption{(a) Speedup comparison, and (b) logarithmic execution times, both for the Shallow water equations, using pure SkelGIS and MSL with the SkelGIS data structure.}
\label{fig:perfs}
\end{figure*}

One can notice that execution times using MSL, which itself uses the SkelGIS distributed data structures, are improved compared to a pure SkelGIS parallelization. Moreover, the scalability stays close to the scalability of SkelGIS. Thus, using a component-based runtime, performance of the back-end code is still HPC-oriented and does not introduce overheads.

\fix{more figures, more conclusions}

### 2015-08-08 WOLFHPC15/article/introduction.tex
fix all helene 08_08

Many solutions to automatically optimize and parallelize partial differential equations (PDEs) solvers have been proposed in the literature, either as specific frameworks~\cite{petsc-efficient,Trilinos-Overview,CPE:CPE3494} or as DSLs~\cite{spaaTangCKLL11,citeulike12258902,Giles2011,DeVito2011LDS}. Most of the time, DSLs focus on the parallelization and the optimization of a single numerical computation, also called a stencil kernel.
However, a real case numerical simulation (such as fluid dynamics, magnetohydrodynamics, molecular synamics etc.) is most of the time not composed of a single stencil kernel, but of a set of stencil kernels and a set of additional local auxiliary computations. The DSL proposed in this paper, called MSL, is a DSL to generate a parallel component-based structure of an overall numerical simulation, that we call a \emph{multi-stencil program}. As it will be detailed in the related work, this is complementary to the optimization and parallelization of a single stencil kernel.

>>>>

Many solutions to automatically optimize and parallelize partial differential equations (PDEs) codes have been proposed in the literature, either as specific frameworks~\cite{CPE:CPE3494} or as DSLs~\cite{spaaTangCKLL11,citeulike12258902,Giles2011,DeVito2011LDS}. Most of the time, DSLs focus on the parallelization and the optimization of a single numerical computation, also called a stencil kernel.
However, a real case numerical simulation (such as fluid dynamics, magnetohydrodynamics, molecular synamics etc.) is most of the time not composed of a single stencil kernel, but of a set of stencil kernels and a set of additional local auxiliary computations. The DSL proposed in this paper, called MSL, is a DSL to generate a parallel component-based structure of an overall numerical simulation, that we call a \emph{multi-stencil program}. As it will be detailed in the related work, this is complementary to the optimization and parallelization of a single stencil kernel.

### 2015-08-08 WOLFHPC15/article/related.tex
fix all helene 08_08

Many specific solutions exist to ease parallel programming of numerical simulations. Solutions such as PETSc~\cite{petsc-efficient} and Trilinos~\cite{Trilinos-Overview} are specific frameworks. They are made to solve many kinds of PDEs using well-known methods and solvers which are efficiently parallelized inside the library. Those frameworks also include solvers for implicit numerical schemes, not addressed in this work. However, using those solutions the user has to use and follow existing tools. In other words, it is not possible for a mathematician or a physician to write his own numerical method if it differs from existing ones. Moreover as PETSc and Trilinos are frameworks they do not propose a domain specific language to define computations. The purpose of those solutions are much more specific than our solution.

Other solutions try to take a step back in the abstraction to propose a way to hide parallel codes for stencils while keeping a sequential programming style. For example PATUS~\cite{citeulike12258902} and Pochoir~\cite{spaaTangCKLL11} are proposed to write stencil codes for general structured meshes, and OP2~\cite{Giles2011} and Liszt~\cite{DeVito2011LDS} for general unstructured meshes. Those solutions can be considered as stencil compilers. Actually they produce optimized (cache tiling etc.) or parallel (CUDA, OpenCL etc.) codes for a single stencil computation.
Recently, domain specific languages for very specific numerical methods such as ExaSlang~\cite{Schmitt:2014:EDL:2691166.2691171} for multigrid solvers, and OPS~\cite{Reguly:2014:ODS:2691166.2691173} for meshes of structured meshes (hybrid mesh) have also been proposed. 
All those solutions are powerful, let the user implement their own stencil codes in a sequential programming style, and produce high performance codes for shared memory architectures and sometimes distributed memory architectures.

Most stencil compilers handle the parallelization or the optimization of a single stencil kernel, considering that it represent the main computation time of numerical simulations. However, many real case numerical simulation are not composed of a single stencil kernel, as explained in Section~\ref{sect:multistencil}, but of more than one stencil computation, involving one or more stencil shapes, and additional local computations. %Solutions like Pochoir or PATUS handle the parallelization or the optimization of a single stencil code, considering that stencil codes represent the main computation time of numerical simulations. 
Liszt~\cite{DeVito2011LDS} is the closest solution to our work as it seems to take into account the entire program and not only stencil kernels.

The work presented in this paper takes place at a different level of numerical simulations and can be seen as complementary to stencil compilers. Actually, the MSL language is agnostic and should work for any type of mesh and any neighborhood shape in stencil codes. The MSC part of the compiler only needs this agnostic description of the overall simulation to generate a parallel pattern of the program, which needs to be filled with computation codes. Thus, depending on the underlying distributed data structure used, those computation codes could be generated by stencil compilers. As a result, MSL and its compiler are not a new contribution to compile optimized stencil codes, but propose another level of parallelization and a different approach of back-end solution.

>>>>

Many specific solutions exist to ease parallel programming of numerical simulations. %Solutions such as PETSc~\cite{petsc-efficient} and Trilinos~\cite{Trilinos-Overview} are specific frameworks. They are made to solve many kinds of PDEs using well-known methods and solvers which are efficiently parallelized inside the library. Those frameworks also include solvers for implicit numerical schemes, not addressed in this work. However, using those solutions the user has to use and follow existing tools. In other words, it is not possible for a mathematician or a physician to write his own numerical method if it differs from existing ones. Moreover as PETSc and Trilinos are frameworks they do not propose a domain specific language to define computations. The purpose of those solutions are much more specific than our solution.
The closest solutions to MSL are DSLs proposing to generate parallel codes from a sequential imperative stencil description. For example PATUS~\cite{citeulike12258902} and Pochoir~\cite{spaaTangCKLL11} propose a simple description to write stencil codes for general structured meshes, and OP2~\cite{Giles2011} and Liszt~\cite{DeVito2011LDS} for general unstructured meshes. Those solutions can be considered as stencil compilers. Actually they produce optimized (cache tiling etc.) or parallel (CUDA, OpenCL etc.) codes for a single stencil computation.
Recently, domain specific languages for very specific numerical methods such as ExaSlang~\cite{Schmitt:2014:EDL:2691166.2691171} for multigrid solvers, and OPS~\cite{Reguly:2014:ODS:2691166.2691173} for meshes of meshes (hybrid mesh) have also been proposed. 
All those solutions let the user implement their own stencil codes in a sequential programming style, and produce high performance codes for shared memory architectures and sometimes distributed memory architectures. 
Most stencil compilers, though, handle the parallelization or the optimization of a single stencil kernel, considering that it represents the main computation time of numerical simulations. However, most real case numerical simulation are composed of more than one stencil kernel, involving one or more stencil shapes, and additional local kernels, as explained in Section~\ref{sect:multistencil}.

Other solutions, as SkelGIS library~\cite{CPE:CPE3494} or Global Arrays~\cite{Nieplocha:2006:AAP:1125980.1125985} propose more or less specialized distributed data structures (specific meshes for SkelGIS, general arrays for Global Arrays), which simplifies or totally hide communications and synchronizations between processes.

The work presented in this paper takes place at a different abstraction level of numerical simulations and can be seen as complementary to stencil compilers or distributed data structures. Actually, the MSL language is agnostic and should work for any type of mesh and any neighborhood shape in stencil codes. The MSCAC compiler only needs this agnostic description of the overall simulation to generate a parallel pattern of the program. This parallel pattern then needs to be filled with computation codes.
%. Thus, depending on the underlying distributed data structure used, those computation codes 
Computation codes could be generated by stencil compilers or could be written using interfaces of distributed data structures (as it is done using SkelGIS in this paper). As a result, MSL is not a new contribution to stencil compilers, or to distributed data structures, but proposes a coarse-grain automatic parallelization of an overall multi-stencil program.
Finally, MSL is a first case study to show that component-based runtimes could improve code-reuse, separation of concerns and thus maintainability of DSLs.
